---
title: "Untitled"
output: html_document
---

```{r, message=FALSE}
library(here)
library(fs)
library(SummarizedExperiment)
library(tidyverse)
```


Load CPTAC3 GBM confirmatory study design (sample names etc)
```{r}
all_samples_tbl = read_tsv(here('resouces/study_design/cptac3_gbm_confirmatory_all_samples.tsv'))
all_samples_tbl %>% head()
```


## DNA methyl. mircoarray annotation
Read EPIC methylation array hg38 manifest and annotation from <https://zwdzwd.github.io/InfiniumAnnotation>
```{r}
# Details about the probe design
EPIC_manifest = readRDS('annotations/EPIC.hg38.manifest.rds')
# Annotation
EPIC_anno = readRDS('annotations/EPIC.hg38.manifest.gencode.v22.rds')
```




## Create sample metadata
The sample table is originally from `/diskmnt/Projects/cptac_scratch/CPTAC3_analysis/Methylation_hg38/batch1219/GBM_sample_batch1221.txt`. But we need the sample naming for the confirmatory cohort
```{r}
read_tsv(
    '/data/dna_methyl/batch1219/GBM_sample_batch1221.txt',
    col_names = c("sample_name", "case_id", "disease", "uuid")
) %>%
    head()
```


Additional information was added from GDC catalog and saved to `gbm_confirm_dna_methyl_gdc_catalog.tsv.gz`
```{r}
samples_tbl = read_tsv('gbm_confirm_dna_methyl_gdc_catalog.tsv.gz') %>%
    rename(case_id = case, aliquot_id = aliquot) %>%
    mutate(
        channel = str_c("channel", str_to_lower(result_type), "uuid", sep = "_"),
        sample_id = str_replace_all(sample_id, glue::glue(",{case_id}"), ",")
    ) %>%
    pivot_wider(
        id_cols = c("case_id", "aliquot_id", "sample_id", "short_sample_type"),
        names_from = "channel", 
        values_from = "UUID"
    )

samples_tbl %>% head()
```

We use sample id to map to the confirmatory sample naming, though some sample ids are not 1-to-1 mappings:
```{r}
samples_tbl$sample_id %>% setdiff(all_samples_tbl$parent_sample_ids)
```

Those IDs above requires manual checks. Create sample metadata
```{r}
meta_tbl = samples_tbl %>%
    mutate(proteome_parent_sample_ids = case_when(
        sample_id == "C3L-04081-01" ~ "C3L-04081-01,-02,-03",
        sample_id == "C3L-07611-02" ~ "C3L-07611-02,-03",
        sample_id == "C3N-03026-01" ~ "C3N-03026-01,-02,-04",
        sample_id == "C3N-03446-02" ~ "C3N-03446-02,-03,-05",
        TRUE ~ sample_id
    )) %>%
    rename(dna_methyl_parent_sample_ids = sample_id) %>%
    left_join(
        all_samples_tbl,
        by = c("case_id", "proteome_parent_sample_ids" = "parent_sample_ids")
    ) %>%
    # Assume the no sample is run twice
    mutate(unique_run_id = preferred_sample_name) %>%
    select(unique_run_id, preferred_sample_name, case_id, sample_type, is_oct, tumor_occurrence_sequence, 
           aliquot_id, dna_methyl_parent_sample_ids, proteome_parent_sample_ids,
           ends_with('uuid'))

meta_tbl
```

Make sure no sample is duplicated
```{r}
stopifnot(meta_tbl %>% count(preferred_sample_name) %>% filter(n > 1) %>% nrow() == 0)
```



## Collect DNA methyl. beta values
We will use hg19 beta values as they are not duplicated.
```{r}
pth_tbl = tibble(hg19_pth = fs::dir_ls('/data/dna_methyl/batch1219/Processed/', glob = '*.csv.zst')) %>%
    mutate(filename = fs::path_file(hg19_pth)) %>%
    extract(
        filename, 
        c("case_id", "aliquot_id"), 
        "^([-\\S]+)\\.\\S+\\.(\\S+).csv.zst$", 
    )

pth_tbl %>% head()
```

Efficient probe value reader
```{r}
read_only_beta_vals = function(zstd_pth) {
    read_csv(
        pipe(paste("zstdcat", zstd_pth)),
        col_types = cols_only(
            Locus = col_character(),
            Beta = col_double()
        )
    )
}
```

```{r}
demo_beta_tbl = read_only_beta_vals(pth_tbl$hg19_pth[[1]])
demo_beta_tbl %>% head()
```

Read all samples
```{r}
beta_tbls = pth_tbl$hg19_pth %>%
    set_names(pth_tbl$aliquot_id) %>%
    map(read_only_beta_vals)
```

Make sure all the samples have the same order of probes
```{r}
stopifnot(every(
    beta_tbls,
    ~ identical(.x$Locus, beta_tbls[[1]]$Locus)
))
```

Combine all values
```{r}
mat = beta_tbls %>%
    map_dfc(~ .x$Beta) %>%
    as.matrix()

rownames(mat) = beta_tbls[[1]]$Locus

# re-order based on the metadata
mat = mat[, meta_tbl$aliquot_id]
# rename the samples
colnames(mat) = meta_tbl$unique_run_id
```

Only keep the probes that passed the filters and are included in the experiment
```{r}
selected_probe_ids = intersect(
    beta_tbls[[1]]$Locus, 
    EPIC_manifest %>% subset(!MASK_general) %>% names()
)
length(selected_probe_ids)
```

```{r}
selected_probes_anno_gr = EPIC_anno[selected_probe_ids, ]
selected_probes_anno_gr
```



## Detection p-values
```{r}
pval_mat = read_csv('/data/dna_methyl/batch1219/Processed/Probewise_pValues.csv') %>%
    column_to_rownames('...1') %>%
    as.matrix()

# re-order based on the metadata
pval_mat = pval_mat[, meta_tbl$aliquot_id]
# rename the samples
colnames(pval_mat) = meta_tbl$unique_run_id
```



## Create the SummarizedExperiment object
```{r}
methyl_se = SummarizedExperiment(
    rowRanges = selected_probes_anno_gr,
    colData = meta_tbl %>% column_to_rownames('unique_run_id'),
    assays = list(
        beta_val=mat[selected_probe_ids, ],
        detect_pval=pval_mat[selected_probe_ids, ]
    ),
    metadata = list(
        cohort = 'CPTAC3 GBM confirmatory cohort',
        description = 'Beta values and probe-wise detection p-values of Illumina EPIC DNA methylation microarry',
        pipeline = str_c(
            'Based on the outputs from Ding Lab CPTAC methylation pipeline v1.1 ',
            'at https://github.com/ding-lab/cptac_methylation. ',
            'Probes were further filtered, keeping MASK_general == FALSE probes based on the annotation ',
            'from https://zhouserver.research.chop.edu/InfiniumAnnotation/20180909/EPIC/EPIC.hg38.manifest.rds. ',
            'More information about the annotation ',
            'at https://zwdzwd.github.io/InfiniumAnnotation (DOI: 10.1093/nar/gkw967)'
        ),
        annotation = str_c(
            'Based on GENCODE v22 ',
            'from https://zhouserver.research.chop.edu/InfiniumAnnotation/20180909/EPIC/EPIC.hg38.manifest.gencode.v22.rds'
        )
    )
)

methyl_se
```

```{r}
saveRDS(methyl_se, 'gbm_confirm_dna_methyl.rds')
```

